---
title: "STA304 Paper #1"
subtitle: "My subtitle if needed"
author: "Ziyao Han"
thanks: "Code and data are available at: https://github.com/HanFrank/STA304-Paper-1"
date: "February 6, 2022"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output: 
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(knitr)
```

# Introduction

First Paragraph is going to be motivational and broad.

Second Paragraph is about what was done and what was found.

Third paragraph is about implications.

Final Paragraph about the remainder of the paper.

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by [@citeTidy] If we were interested in baseball data then could be useful. [@citebookdown] also [@citeknitr] data [@citeodt] 



# Data

Paragraph or two introducing the dataset broadly.

Then show an extract of the dataset (Table \@ref(tab:extract)).

```{r extract, echo = FALSE, warning = FALSE, message = FALSE}
monthly_shelter_usage <- read_csv(here::here("inputs/data/filler_data.csv"))

monthly_shelter_usage %>%
  select(population_group, returned_from_housing) %>%
  slice(1:10) %>%
  kable(
    caption = "First ten rows of a dataset that shows shelter usage",
    col.names = c("Population Group", "Number returned from housing"),
    digits = 1,
    booktabs = TRUE,
    linesep=""
  )
```

Paragraph or two more about it

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}
# This needs to be about the random data that I sampled?
```

Talk way more about it. 

# Discussion

## First discussion point

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References


